Transfer of Learned Opponent Models in Zero Sum Games

### 2020-01-22 Report/draft_report_v1.tex
report changes to comp modelling section

Being able to transfer previously acquired knowledge to a new domain is one of the hallmarks of human intelligence (Taylor, 2008) . Humans are naturally endowed with the ability to extract relevant features from a situation, identify the presence of these features in a novel setting and use previously acquired knowledge to adapt to previously unseen challenges using acquired knowledge. More formally, Perkins (1992) defines transfer of learning as the application of skills, knowledge, and/or attitudes that were learned in one situation to another learning situation. This typically human skill has so far eluded modern AI agents. Deep neural networks for instance can do very well on image recognition tasks and can even reach super-human performance levels on video and strategic board games. Yet they struggle to learn as fast or as efficiently as humans do, and more importantly they have a very limited ability to generalize and transfer knowledge to new domains. Lake, Ullman, Tenenbaum, and Gershman (2017) argue that human learning transfer abilities take advantage of important cognitive building blocks such as an abstract representation of concepts underlying tasks and compositionally structured causal models of the environment.

@@

In most of the experimental paradigms we saw, participants were playing other human players either one on one or as a group. In this case, players are learning about the game but so are their opponents. While the approach of matching groups of human players repeatedly to play economic games has ecological validity in recreating environments where social decision making can be studied, it complicates the detection and modelling of strategies used by participants. It is harder to focus on an individual and how her strategies are changing and adapting to the opponent's play if we cannot experimentally control the behaviour of the opponent. It is also difficult to disentangle the process of learning about the opponent from that of learning about the game structure and payoffs. One potential solution is to replace the human opponent with a computerised agent, allowing researchers to manipulate the opponent's behaviour and observe how the human participants adapt their behaviour. For instance, Spiliopoulos (2013) made humans play constant sum games against 3 computer opponents programmed to take advantage of known patterns in human play such as imperfect randomization and heuristics use and found that human participants do adapt to the opponent they are facing. Shachat \& Swarthout (2004) made human participants face computer opponents playing various mixed strategies in a zero-sum asymmetric matching pennies game. They found that the players changed their strategies towards exploiting the deviations from the Mixed Strategy Nash Equilibrium (MSNE), and that this exploitation was very likely if the deviation from the MSNE play was high.

Still, one of the limitations in studies of how humans adapt to computerised opponents is that they have mostly looked at the ability of players to detect and exploit action-based learning rules. They either used strategies that did not take into account the human participant's histpry of play (as in the case of pre-determined distribution on actions), or strategies that only depended on the player's patterns of past plays, irrespective of the computer play. In other words These strategies did not necessarily take into account the computer's own past plays, or the degree of strategic reasoning of the human player about the computer. Their play was not ``human-like'' in the sense that humans are not very good at playing specific mixed strategies with any precision of at detecting patterns from long sequences of past play due to cognitive constraints. It is therefore important to have agents that ``play like humans'', and one way of achieving that is to embed theses agents with human-like theory of mind abilities based on limited steps of recursive reasoning. Simon (1972) explains that humans have limited cognitive capacities and as such cannot be expected to solve computationally intractable problems such as finding Nash equilibria. Instead, they will try to `satisfice' by choosing a strategy that is adequate in a simplified model of the environment, rather than an optimal one. This concept finds its natural application in `level-k' theory, first adopted by Stahl \& Wilson (1995). It posits that deviations from Nash equilibrium solutions in simple games are explained by the fact that humans have a heterogenous degree of strategic sophistication. At the bottom of the ladder, level-0 players are non-strategic and play either randomly or use a salient strategy in the game environment (Arad \& Rubinstein, 2012). Level-1 players are next up the ladder of strategic sophistication and will assume all their opponents belong to the level-0 category and as such will best respond to them given this assumption. Likewise, a level-2 player will choose actions that are the best response given the belief that all opponents are exactly one level below, and so on.

Another potential issue in the experimental paradigms used in the learning transfer literature is the use of social dilemmas and cooperative/competitive settings that may invoke confounding factors in learning and transfer. Social dilemma and coordination games crystallise the conflict between competing with others out of self-interest or cooperating to reach a socially optimal outcome. Playing these games repeatedly implicates important social constructs such as reputation building, trust and other individual psychological attributes such as cooperativeness and inequity aversion. Coordination games also test the ability of choosing the safe self-interested choice compared to the risky cooperative choice and may depend on similar latent factors. As such, these confounding factors may impede teasing out the process of learning an opponent model and that of whether/how these models are transferred. By contrast, purely competitive settings and zero sum games in particular would be more appropriate. They do not incentivize any cooperation, since one player's gain is necessarily the opponent's loss, and are agnostic to trust mechanisms or reputation building. Also choosing games with no pure strategy Nash Equilibrium shift the focus away from learning a normatively optimal action to that of reasoning dynamically about the opponent which is more amenable to modelling learning. Games with unique mixed strategy Nash equilibrium leading to nil average rewards facilitate inferring whether participants have learned to exploit the non-random play of the opponent.

To address some of these limitations in the existing literature on learning transfer outlined above, we propose to explore opponent modelling and its transfer with the use of computer agents possessing human-like theory of mind abilities with limited degrees of iterated reasoning. The agents will have a fixed strategy played stochastically and will either be level-1 or level-2 behavioral rule, mimicking human theory of mind abilities and the limited recursion depth they exhibit (Goodie et al., 2012). We also use a set of purely competitive zero-sum games that incentivise reasoning about the opponent without introducing social confounding factors. We measure transfer of learning about the opponent strategy between games with varying degrees of similarity. The first two games we use are identical except for action labels. In one experiment, the third game is strategically similar to the first two but descriptively different, while in a second experiment, we introduce a third game that is dissimilar to the first two in terms of payoff matrix and strategic structure while retaining, like all other games, a unique mixed strategy Nash equilibrium of random action with 0 expected reward against a Nash player.

Hence, we ran a total of two experiments where human participants played against computer opponents a total of 3 different games. In the following sections, we wwill describe first the methods used in the first experiments and look at the results, then we will do the same for the second experiment.

@@

To gain more insight into how participants played the games against the computer opponents, we compared multiple models of strategies the players may have been using to learn how to beat the opponent. As a base model, we assume play is random, and each potential action is chosen with equal probability. Note that this corresponds to the Nash equilibrium strategy. Next, we include for comparison purposes a few other basic models of behavior, such as Q-learning on action choices, as well as a more sophisticatd Q-learning model incorporating a state space consisting of last round play. The Bayesian approach is represented by models whereby agents would update an initial belief over the opponent's strategy consisting of various level-k strategies (resp. 0,1 and 2). We fitted two models in which agents best respond to the opponent model. In the first, the beliefs about the opponent's adopted strategy are transfered across games, and another in which no transfer of opponent models is allowed.

Next, we fitted a self-tuning EWA model (Ho et al., 2004) both on action choices without states and with states consisting of lat round play. The self-tuning EWA model generalizes both reinforcement learning as well as fictitious play models and has been proven to do well than equilibrium models in various game situations. Finally, we also included a model-based reinforcement learning approach. whereby the agent learns the transition probabilities between states ans uses that to choose the best action, given the current state and it's belief about what the new state will be.

>>>>

Being able to transfer previously acquired knowledge to a new domain is one of the hallmarks of human intelligence. Humans are naturally endowed with the ability to extract relevant features from a situation, identify the presence of these features in a novel setting and use previously acquired knowledge to adapt to previously unseen challenges using acquired knowledge. More formally, Perkins (1992) defines transfer of learning as the application of skills, knowledge, and/or attitudes that were learned in one situation to another learning situation. This typically human skill has so far eluded modern AI agents. Deep neural networks for instance can do very well on image recognition tasks and can even reach super-human performance levels on video and strategic board games. Yet they struggle to learn as fast or as efficiently as humans do, and more importantly they have a very limited ability to generalize and transfer knowledge to new domains. Lake, Ullman, Tenenbaum, and Gershman (2017) argue that human learning transfer abilities take advantage of important cognitive building blocks such as an abstract representation of concepts underlying tasks and compositionally structured causal models of the environment.

@@

Still, one of the commonalities in studies of how humans adapt to computerised opponents is that they have mostly looked at the ability of players to detect and exploit action-based learning rules. The strategies implemented by the computer opponents had a style of play that was not ``human-like'' in the sense that humans are not very good at playing specific mixed strategies with any precision of at detecting patterns from long sequences of past play due to cognitive constraints. It is therefore important to have agents that ``play like humans'', and one way of achieving that is to embed theses agents with human-like theory of mind abilities based on limited steps of recursive reasoning. Simon (1972) explains that humans have limited cognitive capacities and as such cannot be expected to solve computationally intractable problems such as finding Nash equilibria. Instead, they will try to `satisfice' by choosing a strategy that is adequate in a simplified model of the environment, rather than an optimal one. This concept finds its natural application in `level-k' theory, first adopted by Stahl \& Wilson (1995). It posits that deviations from Nash equilibrium solutions in simple games are explained by the fact that humans have a heterogenous degree of strategic sophistication. At the bottom of the ladder, level-0 players are non-strategic and play either randomly or use a salient strategy in the game environment (Arad \& Rubinstein, 2012). Level-1 players are next up the ladder of strategic sophistication and will assume all their opponents belong to the level-0 category and as such will best respond to them given this assumption. Likewise, a level-2 player will choose actions that are the best response given the belief that all opponents are exactly one level below, and so on.

In this study, we propose to explore opponent modelling and its transfer with the use of computer agents possessing human-like theory of mind abilities with limited degrees of iterated reasoning. The agents will have a fixed strategy played stochastically and will either be level-1 or level-2 behavioral rule, mimicking human theory of mind abilities and the limited recursion depth they exhibit (Goodie et al., 2012).

Our choice of using computer oponents sinstead of matching groups of participants makes it easier to disentangle the process of learning about the opponent from that of learning about the game structure and payoffs. When playing against other human opponents,players are learning about the game and the strategy of the opponent simultaneously. Thus, it is harder to focus on an individual and how her strategies are changing and adapting to the opponent's play if we cannot experimentally control the behaviour of the opponent. The use of computer opponents to elicit learning behavior has been explored in the literature with encouraging results. For instance, Spiliopoulos (2013) made humans play constant sum games against 3 computer opponents programmed to take advantage of known patterns in human play such as imperfect randomization and heuristics use and found that human participants do adapt to the opponent they are facing. Shachat \& Swarthout (2004) made human participants face computer opponents playing various mixed strategies in a zero-sum asymmetric matching pennies game. They found that the players changed their strategies towards exploiting the deviations from the Mixed Strategy Nash Equilibrium (MSNE), and that this exploitation was very likely if the deviation from the MSNE play was high.

We chose to use a set of purely competitive zero-sum games that incentivise reasoning about the opponent without introducing social confounding factors. We avoid social dilemma and coordination games that crystallise the conflict between competing with others out of self-interest or cooperating to reach a socially optimal outcome. Playing these games repeatedly implicates important social constructs such as reputation building, trust and other individual psychological attributes such as cooperativeness and inequity aversion. Coordination games also test the ability of choosing the safe self-interested choice compared to the risky cooperative choice and may depend on similar latent factors. As such, these confounding factors may impede teasing out the process of learning an opponent model and that of whether/how these models are transferred. By contrast, purely competitive settings and zero sum games in particular would be more appropriate. They do not incentivize any cooperation, since one player's gain is necessarily the opponent's loss, and are agnostic to trust mechanisms or reputation building. Also choosing games with no pure strategy Nash Equilibrium shift the focus away from learning a normatively optimal action to that of reasoning dynamically about the opponent which is more amenable to modelling learning. Games with unique mixed strategy Nash equilibrium leading to nil average rewards facilitate inferring whether participants have learned to exploit the non-random play of the opponent.

We measure transfer of learning about the opponent strategy between games with varying degrees of similarity. The first two games we use are identical except for action labels. In one experiment, the third game is strategically similar to the first two but descriptively different, while in a second experiment, we introduce a third game that is dissimilar to the first two in terms of payoff matrix and strategic structure while retaining, like all other games, a unique mixed strategy Nash equilibrium of random action with a fixed expected reward against a Nash player.

Hence, we ran a total of two experiments where human participants played 3 different games against either 1 (first experiment ) or 2 (second experiment) computer opponents. In the following sections, we will describe each experiment's methodology and results separately, then discuss the findings before using computational modelling to shed light on some of the results.

@@

To gain more insight into how participants played the games against the computer opponents, we compared multiple models of strategies the players may have been using to learn how to beat the opponent. As a base model, we assume play is random, and each potential action is chosen with equal probability. Note that this corresponds to the Nash equilibrium strategy.

Next, we include for comparison purposes a simple model-free reinforcement learning algorithm, that reinforces actions that have led to positive rewards, and conversely lowers the likelihood of choosing actions that led to a negative reward, irrespective of any state. We will use a simple temporal difference learning update rule:

\[ V_{t+1}(a) = V_{t}(a) + \alpha*( R_{t}  - V_{t}(a)) \]

Where \$ V\_\{t\}(a) \$ is the vlue associated with action \(a\) at time \(t\), \(\alpha\) is the learning rate and \(R_{t}\) the reward at time t. Actions are chosen based on action values using a softmax choice rule.

We extend this very simple model by adding a state space that consists of last round human and agent play. This is akin to using a Q-learning algorithm ({\textbf{???}}). The update rule becomes:

\[ Q^{new}(s_{t},a_{t}) = Q(s_{t},a_{t}) + \alpha*( R_{t}  + \gamma*\underset{a}{\max}Q(s_{t+1},a) - Q(s_{t},a_{t}) ) \]

Where \(Q(s_{t},a_{t})\) is the value of taking action \(a\) when in state \(s\) at time \(t\), \(\gamma\) is the discount rate applied to future rewards. For instance, \(Q(RS,P)\) denotes the value of taking action ``Paper'' this round if the player's last action was ``Rock'' and the opponent played ``Scissors''. This is a much richer model allowing the players to compute the values of actions conditional on past play.

\hypertarget{ewa-models}{%
\subsubsection{EWA Models}\label{ewa-models}}

Next, we use a self-tuning Experience Weighted attraction model ({\textbf{???}}). EWA models are model of behavior in repeated games, that nest two seemingly different approaches, namely reinforcement learning and belief learning. Belief based models are based on the assumption that players keep track of the frequency of past plays and best respond to that. In contrast, reinforcement learning does not take into account beliefs about other players, but is such that an action followed by a positive reward is more likely to be repeated than an action followed by a negative reward. The self-tunign EWA model has been shown to perform better than both these nested models in multiple repeated gmes and has the advantage of having only one free parameter, the inverse temperature in the softmax choice function.

Lt's define some notation in order to write the pdate rule of the self-tuning EWA model. For player i, there are \(m_{i}\) strategies, denoted \(s_{i}^{j}\) (i.e player i's strategy number j). Strategies actually played by i in period t, are denoted \(s_{i}(t)\), while the opponent's strategy at time t is denoted \(s_{-i}(t)\). After playing \(s_{i}^{j}\) at time \(t\), player i payoff is denoted \(\pi_{i}(s_{i}^{j},s_{-i}(t))\), and the actual payoff the player received is \(\pi_{i}(t)\).

The EWA model is based on updating ``Attractions'' for each action over time. For instance, the attraction of strategy \(j\) to player \(i\) at time \(t\) is written \(A_{i}^{j}(t)\). Future action choice probabilities are based on these attractions using the softmax playing rule:

\[P_{i}^{j}(t+1) = \frac{e^{\lambda . A_{i}^{j}(t)}}{\sum_{k=1}^{m_{i}}e^{\lambda . A_{i}^{k}(t)}} \]

The attractions are updated over every time step \(t\) using the followng update rule :

\[ A_{i}^{j}(t) =  \frac{\phi . N(t-1) . A_{i}^{j}(t-1) + [ \delta + (1-\delta) . I(s_{i}^{j},s_{i}(t))] . \pi_{i}(s_{i}^{j},s_{-i}(t)) } {\phi . N(t-1) + 1} \]

Here, I(x,y) is the indicator function equla to 1 if \$ x = y\$ and 0 otherwise. A simple way to think about this update rule is that attractions are multipled by a parameter that represents experince (\(N(t-1)\)) which is itself decaying by a wight \(\phi\). The result is added to either the payoff received (when the indicator function is 0), or to \(\delta\) times the foregone payoff (when indicator function is 1). We can see that setting \(\delta = 0\) leads to reinforcement of past actions, while positive and high delta parameters make the update rule take into account foregone payoffs, which is similar to weighted fictitous play ({\textbf{???}}).
While the assumption in expanding the update rule above is that \(\phi\) and \(\delta\) are free parameters ({\textbf{???}}), the self-tuning aspect of the model comes from the fact that these are now self-tuned using the formulas expanded in ({\textbf{???}}).

\hypertarget{tom-models}{%
\subsubsection{ToM models}\label{tom-models}}

In this set of models, we assume that participants have a belief that the opponent is a level-k agent, with uniform probability of the level k, and use evidence of past play to update their beliefs in a Bayesian way about the true value of k. We use values of k in \({0,1,2}\). We distinguish between multiple ToM models based on their ability to keep what was learned about the opponent in memory and hence facilitate transfer. In a No-Between-Transfer (NBT) model, players have no memory of what was learned about the opponent and start every new game assuming each level-k has equal probability. In the context of Experiment 2 where players face both opponents, this model assumes that participants transfer learning within the same game, from the first to the second intraction with the opponent, but are not able to transfer that learning to new games.

Conversely, In a Between-Transfer model (BT), players are assumed to keep in memory what was learned about the type of opponent faced (vector of probabilities of level-k) and use that at the beginning of each new game. In the context of experiment 2, we still assume that within game transfer is still present (from first to second interaction).

In experiment two, on top of these two models, we fit two other models: One in which we assume that players don't distinguish between the two opponents faced, so the opponent model vector is never updated (Naive). And finally, we also fit a model in which all stages of the game and all new games start with a fresh uniforme probability of level-k opponent (NT), so no within or between opponent model learning transfer.

In summary, we fit two ToM models for experiment 1-play (NBT and BT), and 4 ToM models for experiment 2 ( NBT, BT, NT and Naive)
